Machine Learning can Advance Theory Formation in the Social Sciences

### 2023-11-24 presentation_copenhagen2023.Rmd
update 2023-11-24 10:41:01.673245

**Theory**: A model that describes the nature of relations between several phenomena in sufficient detail that it allows for the derivation of quantifiable hypotheses that can be subject to (severe) testing.

* Most social scientific theories are:
    + insufficiently precise
    + too flexible
* Compare to @walasekEvolutionaryModelSensitive2022

Why do we need theory?

* Useful as an instrument of cumulative knowledge acquisition
* Communicating social scientific research to its consumers

@@

De Groot’s empirical cycle: a model of knowledge production through scientific research [@degrootMethodologieGrondslagenVan1961]

@@

* automated model building
* checks and balances to prevent overfitting
* maximize predictive performance and generalizability of results

@@

First step in *Theory Construction Methods* is identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020]

* Phenomena: stable and general features of the world
* Few quantitative methods for phenomena detection (aside from expert opinion)

**Text mining** may be suitable for phenomena detection [@vanlissaMappingPhenomenaRelevant2021]

* Unsupervised learning method
* Clusters in keywords and abstracts

@@

* Assumption for causal interpretation
    + Multicollinearity
* Good theory incorporates most important causes
    + Cannot assess the relative utility across studies
* Include potential predictors from various theories, and undertheorized factors

@@

Some machine learning methods incorporate theoretical elements

* E.g.: assumption that development follows a latent growth curve (LGC).
* SEM forests [@brandmaierTheoryguidedExplorationStructural2016]
* regularized SEM [@jacobucciRegularizedStructuralEquation2016], (Ernst & Peikert 2022)

@@

All of these methods allow for theory guided exploration using machine learning

@@

@@

* E.g., latent class analyses
    + "Which individuals are similar?"
    + Other *unsupervised learning* methods exist
* RI-CLPM and DSEM
    + Rarely explain heterogeneity in within-person effects
    + Regularization
* Tree-based models
    + Group individuals based on predictors to maximize homogeneity of the outcome
    + "Why are these individuals similar?"

@@

When using data models to guide theory formation, generalizability is essential

Checks and balances to curtail *overfitting*

@@

* Variable importance metrics
    + congruent with theoretical assumptions about important predictors?
    + any theoretically important predictors rank low?
    + any undertheorized factors rank high?

* Marginal associations
    + non-linear effects?
    + high importance but flat marginal association?

Comparing predictive performance of simpler parametric model that represents these 'insights' to machine learning model

@@

* abductive formal theory construction (AFTC) framework
    + construct formal theory
    + Generate data
    + Mathematically compare to empirical data
    + like a posterior predictive check in Bayes
    + Discrepancies: amend formal theory

@@

<!-- ## Knowledge gap -->

<!-- Substantial empirical research, but: -->

<!-- * Different (sub)disciplines have approached the topic in disparate ways [@riedigerEmotionRegulationAdolescence2014] -->
<!-- * No consistent terminology [@bariolaChildAdolescentEmotion2011] -->
<!-- * No conceptual frameworks [@stifterEmotionRegulation2019] -->
<!-- * No overarching theoretical framework [@bussTheoriesEmotionalDevelopment2019] -->

<!-- Knowledge must be consolidated into overarching theory -->

@@

@@

@@

## Characteristics of formal theory

According to Meehl:

@@

## Formal theory

* Characteristics 1-2 can be described by a causal network (DAG)
* Characteristics 3-6 refer to the nature of associations within that network

Augmented causal networks (ACN, see also Crielaard et al., 2022)

@@

@@

* Using TMSR, based on what previous authors have written
* Using causal discovery, based on large panel studies
* Using ESM data to leverage precedence in time
    + Causality requires association, precedence in time, and nonspuriousness

## Future direction 2: Functional form

Previous work: Post-hoc interpretation of marginal associations

Future direction: Use a-priori interpretable machine learning

* Symbolic regression applies deep learning or genetic algorithms to learn a mathematical formula that describes the relationship between an outcome and its predictors

@@

Van Lissa, C. J. (2022). Developmental data science: How machine learning can advance theory formation in Developmental Psychology. Infant and Child Development. https://doi.org/10.1002/icd.2370

@@

##

>>>>

> **Theory**: A (causal?) explanation of phenomena reliably evidenced by data [@bogenSavingPhenomena1988]

* Allows for derivation of testable hypotheses for deductive research

## Characteristics of Good Theory

@meehlAppraisingAmendingTheories1990

Most social scientific theories are [@scheelWhyMostPsychological2022]:

* insufficiently precise
* too flexible
* But cf. @walasekEvolutionaryModelSensitive2022

## Why do we need theory?

* Social science strives to understand, not only predict
* Useful as an instrument of cumulative knowledge acquisition
* Communicating social scientific research to its consumers

@@

De Groot’s empirical cycle

* Model of cumulative knowledge acquisition through scientific research [@degrootMethodologieGrondslagenVan1961]

@@

* Automated model building
* Checks and balances to prevent overfitting
* Maximize predictive performance and generalizability of results

@@

First step in *Theory Construction Methods* is identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020]
*
* Few quantitative methods for phenomena detection (aside from expert opinion)
* Phenomena: stable and general features of the world
* **Cross-validated patterns of association** may constitute phenomena
* **Text mining** existing literature may reveal what phenomena other authors have identified [@vanlissaMappingPhenomenaRelevant2021]

@@

* Assumption for causal interpretation
    + But mind multicolinearity and collider bias
* Good theory incorporates most important causes
* Include potential predictors from various theories, and undertheorized factors
    + Assess the relative utility across studies
    + Cannot do this across studies

@@

Some methods allow for theory guided exploration using machine learning

When (part of) the theory can be expressed as a structural equation model, and part still remains to be discovered through exploratory research

* E.g., some relevant constructs are latent variables
* E.g., development follows a latent growth curve (LGC)
* SEM forests [@brandmaierTheoryguidedExplorationStructural2016]
* regularized SEM [@jacobucciRegularizedStructuralEquation2016], (Ernst & Peikert 2022)

@@

@@

Call for person-centered research [@laursenPersonCenteredVariableCenteredApproaches2006]

@@

* Latent class analyses
    + "Which individuals are similar?"
    + Other *unsupervised learning* methods exist
* Multilevel models
    + Allow for, but rarely explain heterogeneity
    + Includes RI-CLPM and DSEM
    + Could use regularized SEM or SEM forests
* Tree-based models
    + Group individuals based on predictors to maximize homogeneity of the outcome
    + "Why are these individuals similar?"

@@

When using data models to guide theory formation, generalizability is essential [@haslbeckModelingPsychopathologyData2021]

ML: Checks and balances to curtail *overfitting*

@@

* Start with "interpretable machine learning" metrics [@henningerInterpretableMachineLearning2022]

* Variable importance
    + congruent with theoretical assumptions about important predictors?
    + any theoretically important predictors rank low?
    + any undertheorized factors rank high?
* Marginal associations
    + non-linear effects?
    + high importance but flat marginal association?
* Exemplar-based interpretation

@@

* Abductive Formal Theory Construction (AFTC) framework
    + construct formal theory
    + Generate data
    + Mathematically compare to empirical data
    + like a posterior predictive check in Bayes
    + Discrepancies: amend formal theory
* "Light" version of AFTC: Comparing predictive performance of simpler parametric model based on "interpretation" to more flexible machine learning model

@@

## Knowledge gap

Substantial empirical research, but:

* Different (sub)disciplines have approached the topic in disparate ways [@riedigerEmotionRegulationAdolescence2014]
* No consistent terminology [@bariolaChildAdolescentEmotion2011]
* No conceptual frameworks [@stifterEmotionRegulation2019]
* No overarching theoretical framework [@bussTheoriesEmotionalDevelopment2019]

Knowledge must be consolidated into overarching theory

@@

## Next Steps

Rasoul Norouzi is further developing this method

* Goal: Reconstructing a causal loop diagram based on authors' causal claims in literature

Together with Denny Borsboom, applying TMSR v2 to Cooperation Databank

* Using sciBERT vector embeddings and HDBSCAN clustering instead of dictionary

@@

# Example: Predictors of Infection Prevention Behavior

## Context

Prior to vaccines, infection prevention behaviors were the only defense against COVID

Behavioral science could thus inform policy

Existing theories suggest a limited set of relevant predictors

We used ML to cast a wider net and identify important predictors at the individual and country level

* PsyCorona consortium
* 56.072 participants across 28 countries
* Collected between March-May 2020

## Variables

* DV: Self-reported infection prevention behaviors (handwashing, avoiding crowds, isolation)
* IVs: 80 individual-level variables (16 multi-item composites)
* Enriched with 35 country-level variables
    + E.g., political stability, government effectiveness
    + Some matched to date of participation (pandemic severity, government response)

<https://github.com/cjvanlissa/covid19_metadata>

```{r, out.height="100%"}
knitr::include_graphics("presentation/varimp_covid.png")
```

* Injunctive norms most important
    + Consistent with theory of planned behavior
* Coping styles were important:
    + Problem-focused coping, focus on the present/ future, hopefulness
* Five of most important predictors were country-level variables
    + E.g., national healthcare infrastructure

## Notable absences

* Perceived likelihood of infection
    + Contradicts Health Belief Model
* Psychological and material struggle
    + E.g., loneliness, boredom, job insecurity
* No evidence for "intergenerational conflict of interest"
    + Confirmed by preregistered studies

## Strengths

This sample included a representative sub-sample

* Allowed us to estimate generalizability
* Predictive $R^2$ in the representative subsample was even higher than in the total hold-out sample ($R^2_{rep} = .59$, $R^2_{test} = .52$)

@@

## Recall: Characteristics of Good Theory

@meehlAppraisingAmendingTheories1990

@@

## Formal Theory

* 1-2 can be described by a causal network (DAG), 3-6 refer to the nature of associations within that network
* Augmented causal networks (ACN, see Crielaard et al., 2022)

@@

Working on this with Aaron Peikert, Andreas Brandmaier and Felix Schönbrodt

@@

For example, using TMSR, based on what previous authors have written

Working on this with my PhD student, Rasoul Norouzi

## Epistemology

Working on an epistemology of machine learning-informed theory formation

With Denny Borsboom and James Woodward

@@

<font size = 4>
**Van Lissa, C. J. (2022). Developmental data science: How machine learning can advance theory formation in Developmental Psychology. Infant and Child Development. https://doi.org/10.1002/icd.2370**

@@

Douglas, K., Sutton, R. M., Van Lissa, C. J., ... & N Pontus, L. (2023). Identifying Important Individual-and Country-Level Predictors of Conspiracy Theorizing: A Machine Learning Analysis. European Journal of Social Psychology. DOI: 10.1002/ejsp.2968

</font>

## More refs
